---
title: "Question 3: How does the number of people flying between different locations change over time?"
author: "Ren Ern"
date: '2022-03-22'
output:
  html_document:
    theme: darkly
    df_print: paged
---
### Loading essential libraries
##### Leaflet is an open source JavaScript library used to build web mapping applications. Leaflet will be used to plot the coordinates of destinations on the world map.
```{r warning = FALSE, message=FALSE}
library(DBI)
library(RODBC)
library(odbc)
library(dbplyr)
library(leaflet)
library(htmlwidgets)
library(maps)
pacman::p_load(pacman, dplyr, GGally, ggplot2, ggthemes, 
  ggvis, httr, lubridate, plotly, rio, rmarkdown, shiny, 
  stringr, tidyr, psych) 
con <- DBI::dbConnect(odbc::odbc(), "PostgreSQL30")
```






##### A reference connection to the table dfff (as seen in the data cleaning phase) is created.
```{r warning = FALSE, message=FALSE}
db <- tbl(con, "dfff") #taking a reference to the remote database
db
```





##### The flights of 1987, between a time period of October to December, shall be used. The datasets are split based on these respesctive months.
```{r warning = FALSE, message=FALSE}
coors <- db  %>% collect()
Oct <- coors  %>%filter(month(date)=='10') 
Nov <- coors  %>%filter(month(date)=='11') 
Dec <- coors  %>%filter(month(date)=='12') 
```





## October flights destination
```{r warning = FALSE, message=FALSE}
Oct_Map <- leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>% 
  setView(lng = -74.0060, lat = 40.7128, zoom = 10) %>%
# Adding clusterOptions = markerClusterOptions()
  addMarkers(lng = Oct$long, lat = Oct$lat, 
               clusterOptions = 
               markerClusterOptions())
Oct_Map
```






## November flights destination
```{r warning = FALSE, message=FALSE}
Nov_Map <- leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>% 
  setView(lng = -74.0060, lat = 40.7128, zoom = 10) %>%
# Adding clusterOptions = markerClusterOptions()
  addMarkers(lng = Nov$long, lat = Nov$lat, 
               clusterOptions = 
               markerClusterOptions())

Nov_Map
```






## December flights destination
```{r warning = FALSE, message=FALSE}
Dec_Map <- leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>% 
  setView(lng = -74.0060, lat = 40.7128, zoom = 10) %>%
# Adding clusterOptions = markerClusterOptions()
  addMarkers(lng = Dec$long, lat = Dec$lat, 
               clusterOptions = 
               markerClusterOptions())

Dec_Map
```





## Alternative visualization method using ggplot
```{r warning = FALSE, message=FALSE}
library(maps)
library(mapdata)
library(gganimate)
library(gifski)
```





#### Grouping flights by their dates and destination, and adding a frequency value.
```{r warning = FALSE, message=FALSE}
df <- coors %>% count(date, lat, long)
df
```





#### Creating a base map
```{r warning = FALSE, message=FALSE}
state <- map_data("state")

basemap <- ggplot(data=state, aes(x=long, y=lat, fill=region, group=group)) + 
  geom_polygon(color = "black", fill="white") + 
  guides(fill=FALSE) + 
  theme(axis.title.x=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank(),
        axis.title.y=element_blank(), axis.text.y=element_blank(), axis.ticks.y=element_blank()) + 
  ggtitle('U.S. Map with States') 
  
basemap
```





#### Adding points to the map
##### The max range for latitude and longitude values have to be specefied. If not, the map would be very small.
```{r warning = FALSE, message=FALSE}
min_long <- -125
max_long <- -68.01197
min_lat <- 25
max_lat <- 50

map_with_data <- basemap +
  geom_point(data = df, aes(x = long, y = lat, fill = 'black' ,color=n, size=n, group=date))  +
  coord_quickmap(xlim = c(min_long, max_long),  ylim = c(min_lat, max_lat)) 

map_with_data
```





#### The frames of the map is then animated according to the number of distinct days in the dataset, which in this case is 92.
```{r warning = FALSE, message=FALSE}
df %>% summarise(count = n_distinct(date))
map_with_animation <- map_with_data +
  transition_time(date) +
  ggtitle('Date: {frame_time}',
          subtitle = 'Frame {frame} of {nframes}') +
  shadow_mark() +
  scale_color_gradientn(colors = rainbow(7))

animate(map_with_animation, nframes = 92, fps = 2)
anim_save("flight_map.gif", map_with_animation, nframes = 92, fps = 3)
```