---
title: 'Question 2: Do older planes suffer more delays? '
author: "Ren Ern"
date: '2022-03-22'
output:
  html_document:
    theme: darkly
    df_print: paged
---
### Loading essential libraries
```{r, warning = FALSE, message = FALSE}
library(DBI)
library(RODBC)
library(odbc)
library(dbplyr)
pacman::p_load(pacman, dplyr, GGally, ggplot2, ggthemes, 
  ggvis, httr, lubridate, plotly, rio, rmarkdown, shiny, 
  stringr, tidyr, psych) 
library(ggpubr)
con <- DBI::dbConnect(odbc::odbc(), "PostgreSQL30")
```





##### A reference connection to the table dff (as seen in the data cleaning phase) is created.
```{r, warning = FALSE, message = FALSE}
db <- tbl(con, "dff") #taking a reference to the remote database
db
```






## Exploring relationship between age and departure delay
```{r, warning = FALSE, message = FALSE}
age_delay <- db %>% filter(departure_delay>0)%>% 
  mutate(age = year(scheduled_departure) - as.integer(year))%>% 
  group_by(age) %>% 
  summarize(Average_Departure_Delay = mean(departure_delay)) %>% collect()

sp <- ggscatter(age_delay, x = "age", y = "Average_Departure_Delay",
   add = "reg.line",  # Add regressin line
   add.params = list(color = "blue", fill = "lightgray"), # Customize reg. line
   conf.int = TRUE # Add confidence interval
   )
# Add correlation coefficient
sp + stat_cor(method = "pearson", label.x = 3, label.y = 30)
```

##### Generally, older planes do suffer greater departure delays, as seen from the graph. The relationship between two variables is generally considered strong when their r value is larger than 0.7. The r value for this relationship is 0.37, indicating a fairly strong positive correlation.







## Exploring relationship between age and departure delay, with age caterogies 
##### According to the standards of the airline industry, a plane is considered new when it is less of 10 years, standard age when it is more then 10 and less then 20 years old, and old after 20 ##### years+. The planes are categorized into "new", "standard" and "old". The relationship between age and delays are then explored individually for each category.
```{r, warning = FALSE, message = FALSE}
age_cat_delay <- db %>% filter(departure_delay>0)%>% 
  mutate(age = year(scheduled_departure) - as.integer(year))%>% 
  mutate(AgeStatus = case_when(age<10 ~ "New", age<20 & age>=10 ~ "Standard",  age>=20 ~ "Old"))%>% 
  group_by(age, AgeStatus) %>% 
  summarize(Average_Departure_Delay = mean(departure_delay)) %>% collect()

sp <- ggscatter(age_cat_delay, x = "age", y = "Average_Departure_Delay",
   color = "AgeStatus", palette = "jco",
   add = "reg.line", conf.int = TRUE)
sp + stat_cor(aes(color = AgeStatus), label.x = 3)
```

##### Suprisingly, the correlation between delays and age of new aircrafts is the strongest, with an R value of 0.65.
##### For standard aircrafts, the correlation is negative, with an R value of = 0.7.






## Exploring relationship between age and arrival delay
```{r, warning = FALSE, message = FALSE}
age_arrival_delay <- db %>% filter(departure_delay>0)%>% 
  mutate(age = year(scheduled_departure) - as.integer(year))%>% 
  group_by(age) %>% 
  summarize(Average_Arrival_Delay = mean(arrival_delay)) %>% collect()

sp <- ggscatter(age_arrival_delay, x = "age", y = "Average_Arrival_Delay",
   add = "reg.line",  # Add regressin line
   add.params = list(color = "blue", fill = "lightgray"), # Customize reg. line
   conf.int = TRUE # Add confidence interval
   )
# Add correlation coefficient
sp + stat_cor(method = "pearson", label.x = 3, label.y = 30)
```

##### Generally, older planes do suffer greater arrival delays, as seen from the graph. TThe r value for this relationship is 0.5, indicating a strong positive correlation.







## Exploring relationship between age and arrival delay, with age caterogies 
```{r, warning = FALSE, message = FALSE}
age_cat_delay <- db %>% filter(departure_delay>0)%>% 
  mutate(age = year(scheduled_departure) - as.integer(year))%>% 
  mutate(AgeStatus = case_when(age<10 ~ "New",age<20 & age>=10 ~ "Standard",age>=20 ~ "Old"))%>% 
  group_by(age, AgeStatus) %>% 
  summarize(Average_Arrival_Delay = mean(arrival_delay)) %>% collect()

sp <- ggscatter(age_cat_delay, x = "age", y = "Average_Arrival_Delay",
   color = "AgeStatus", palette = "jco",
   add = "reg.line", conf.int = TRUE)
sp + stat_cor(aes(color = AgeStatus), label.x = 3)
```

##### The age of the new aircrafts have a very strong correlation with arrival delays, with an R value of 0.69. This does not hold true for standard and old planes.